\texttt{Supertree construction with}\texttt{\emph{ Incertae sedis}}\texttt{ taxa}

### 2018-03-15 doc/incertae_sedis_latex2.tex
Update doc.

%% LyX 2.2.3 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english]{article}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{geometry}
\geometry{verbose,tmargin=3cm,bmargin=3cm,lmargin=3cm,rmargin=3cm}
\setlength{\parskip}{6pt}
\setlength{\parindent}{0pt}
\usepackage{babel}
\usepackage{verbatim}
\usepackage{url}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[authoryear]{natbib}
\usepackage[unicode=true]
 {hyperref}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% Special footnote code from the package 'stblftnt.sty'
%% Author: Robin Fairbairns -- Last revised Dec 13 1996
\let\SF@@footnote\footnote
\def\footnote{\ifx\protect\@typeset@protect
    \expandafter\SF@@footnote
  \else
    \expandafter\SF@gobble@opt
  \fi
}
\expandafter\def\csname SF@gobble@opt \endcsname{\@ifnextchar[%]
  \SF@gobble@twobracket
  \@gobble
}
\edef\SF@gobble@opt{\noexpand\protect
  \expandafter\noexpand\csname SF@gobble@opt \endcsname}
\def\SF@gobble@twobracket[#1]#2{}
%% A simple dot to overcome graphicx limitations
\newcommand{\lyxdot}{.}

@@

\begin{document}
\global\long\def\taxonomy{\mbox{\ensuremath{\mathbb{T}}}}

@@

\title{\texttt{\emph{Incertae sedis}}\texttt{ taxa in supertree construction}}
\maketitle

@@

The phrase \textquotedblleft \emph{incertae sedis}\textquotedblright ,
which is a Latin phrase meaning ``uncertain seat'', is used in classifications
to indicate that a group is of uncertain taxonomic placement. Such
groups arise in practice when they are classified within a parent
group, such as Fungi, but the precise location within the parent group
is uncertain. Such a taxon would then be attached as a child of the
parent group, but labeled as \emph{incertae sedis}''.

The OpenTree project seeks to build a comprehensive supertree covering
all of life. The approach is to combine information in published phylogenies
with a comprehensive taxonomy that supplies taxon names \citep{redelings2017supertree}.
We describe an extension to our supertree method so that we can resolve
taxonomic uncertainty by using published phylogenies to place \emph{incertae
sedis} taxa. We describe an updated semantics for conflict between
phylogeny edges and taxonomy edges and an updated semantics for assigning
taxonomy names to nodes in the synthetic tree.

The ability to place taxa within the taxonomy without breaking their
containing taxa enables us to include thousands of new taxa that were
previously filtered out to avoid broken taxa. It also enables us to
use phylogenetic trees to update the taxonomy by extending taxa to
include \emph{incertae sedis} taxa placed within them. This makes
substantial progress towards our goal of complete representation of
taxa. It also enables us to include extinct taxa, since many of these
taxa are incertae sedis.

\emph{TODO: add cases}

\paragraph{Formalism}

Formalism. We consider a ranked collection $\phyloinputs=\left\{ T_{1},\ldots,T_{n}\right\} $
of input trees and a single taxonomy tree $\taxonomy$. We have a
set of labels $\mathcal{L}$ that correspond to taxon names. Therefore
taxonomy node has a label, and every label correspond to a unique
taxonomy node.

In contrast, input phylogenies only have labels on their leaf nodes.
Furthermore, the set of labels $\mathcal{L}(i)$ on input tree $T_{i}$
need not include all leaf labels, and in practice always contains
only a relatively small part of $\mathcal{L}$.

We extend this framework by adding a set $\mathcal{I}\subseteq\mathcal{L}$
of labels that have the \emph{incertae sedis }property. The taxonomy
consists of the labels $\mathcal{L}$, the taxonomy tree $\taxonomy$,
and the \emph{incertae sedis} property $\mathcal{I}$ of taxon names.

\section{Details \& Methods}

\subsection{Some examples in OTT\protect\footnote{This paragraph pulled from intro to otcetera/doc/handling-incertae-sedis.pdf}}

The problem is becoming more acute because NCBI is putting a larger
number of taxa into groups that are marked as \textquotedblleft unclassified.\textquotedblright{}
For example, when OTT 2.9 was created NCBI\textquoteright s clasification
of the bird family Sylvidae included a group five genera that were
placed within \textquotedblleft \href{http://www.ncbi.nlm.nih.gov/Taxonomy/Browser/wwwtax.cgi?mode=Undef&id=36270&lvl=3&lin=f&keep=1&srchmode=1&unlock}{unclassified Sylviidae}.\textquotedblright{}
This includes the genus Regulus. Thus in OTT 2.9 Regulus (ott ID =
3599326) is placed inside the family Silvidae (OTT ID = 259942). The
\textquotedblleft unclassified Sylviidae\textquotedblright{} does
not appear in OTT; instead Regulus is flagged as \textquotedblleft unclassified,sibling
higher\textquotedblright{} and all of the species within Regulus are
flagged as \textquotedblleft unclassified inherited.\textquotedblright{}

The taxonomy (as of version 2.9) also contains 685 cases of taxa that
are flagged as both \textquotedblleft unclassified inherited\textquotedblright{}
and \textquotedblleft unclassified\textquotedblright \footnote{based on \texttt{grep unclass.{*}unclassified\textbackslash{}\_inherited
taxonomy.tsv}}.

We would like to stop suppressing (pruning) incertae sedis taxa, so
that groups such as Regulus can appear appear in the synthetic tree.

\begin{comment}
While we get Regulus back, we do not end up placing it in Sylviidae
because Sylviidae is broken. It conflicts with about 20 different
trees.
\end{comment}

\subsection{Taxonomy details}

\textbf{Flags.} The Open Tree\textquoteright s reference taxonomy
is produced by a tool called smasher that notices hints that a taxon
is incertae sedis and labels that taxon with one of five flags\footnote{see \url{https://github.com/OpenTreeOfLife/reference-taxonomy/wiki/Taxon-flags}.This
document is not concerned with the flagging system \emph{per se},
so \textquotedblleft \emph{incertae sedis}\textquotedblright{} will
be used here to refer to all of the flags that denote taxa with uncertain
placement.}. While a taxonomist may intend to use incertae sedis to indicate
a limited number of possible positions for a taxon to go in the taxonomy,
we do not retain any such details.

\textbf{Container nodes.} Some taxonomies have nodes with names like
``Incertae sedis (Bacteria)''. Such nodes are called containers.
They indicate that each child node is an \emph{incertae sedis} child
of the container node's parent. The OTT retains container nodes with
the \emph{was\_container} flag, but seems to have already moved all
of the container's children to the container's parent.

@@

>>>>

%% LyX 2.3.0 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english]{article}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{geometry}
\geometry{verbose,tmargin=3cm,bmargin=3cm,lmargin=3cm,rmargin=3cm}
\setlength{\parskip}{6pt}
\setlength{\parindent}{0pt}
\usepackage{verbatim}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[authoryear]{natbib}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% A simple dot to overcome graphicx limitations
\newcommand{\lyxdot}{.}

@@

\usepackage{babel}
\begin{document}
\global\long\def\taxonomy{\mbox{\ensuremath{\mathbb{T}}}}

@@

\title{\texttt{Supertree construction with}\texttt{\emph{ Incertae sedis}}\texttt{
taxa}}
\maketitle

@@

The phrase \textquotedblleft \emph{incertae sedis}\textquotedblright ,
which is a Latin phrase meaning ``uncertain seat'', indicates that
a taxon is of uncertain placement. Such groups arise in practice when
they are classified within a parent group, such as Fungi, but the
precise location within the parent group is uncertain. Such a taxon
would then be attached as a child of the parent group, but labeled
as \emph{incertae sedis}''.

The OpenTree project seeks to build a comprehensive supertree covering
all of life. The approach is to combine information in published phylogenies
with a comprehensive taxonomy that supplies taxon names \citep{redelings2017supertree}.
We describe an extension to this supertree method so that we can resolve
taxonomic uncertainty by using published phylogenies to place \emph{incertae
sedis} taxa. We describe an updated semantics for conflict between
phylogeny edges and taxonomy edges and an updated semantics for assigning
taxonomy names to nodes in the synthetic tree.

The ability to place taxa within the taxonomy without breaking containing
taxa enables us to include thousands of new taxa that were previously
filtered out to avoid broken taxa. It also enables us to use published
phylogenies to update the taxonomy by extending taxa to include \emph{incertae
sedis} taxa placed within them. This makes substantial progress towards
our goal of complete representation of taxa. It also enables us to
include extinct taxa, since many of these taxa are incertae sedis.

\paragraph{Background}

We briefly describe the supertree algorithm proposed in \citet{redelings2017supertree}.
This supertree algorithm takes as input a ranked list $\phyloinputs=\left\{ T_{1},\ldots,T_{n}\right\} $
of rooted input trees and a single rooted taxonomy tree $\taxonomy$.
The taxonomy tree is considered by ranked below any of the input trees.
Each node $i$ in the taxonomy tree is associated a taxon name $n(i)$
in the set of taxon names $\mathcal{N}$, and each taxon name is corresponds
to only one node of the taxonomy tree. Thus the taxonomy is comprehensive
and contains all taxa. In contrast, input phylogenies only have labels
on their leaf nodes, and generally include a small fraction of the
total number of taxa. Furthermore, taxa in input trees need not correspond
to leaf taxa on the taxonomy tree, and multiple tips may be labeled
with the same taxon name. \citet{redelings2017supertree} described
a method called 'exemplification' that reduces this problem to a simpler
form where all taxon names of input trees (i) correspond to leaf taxa
and (ii) occur only once in an input tree. For the purposes of this
paper, we therefore assume that this is the case.

Given the exemplified input trees and taxonomy, the supertree algorithm
first divides the supertree problem into independent pieces, solves
the problem on these pieces, and then glues the pieces back together.
This is made possible by the assumption that any taxonomy edges that
are not contested by any of the input trees can be assumed to occur
in the supertree. We may then solve separately for what happens on
each side of such an uncontested edge. When the uncontested edge is
attached to the solution below the edge and the solution above the
edge, the complete solution is then obtained.

Supertrees for each subproblem are obtained by greedily constructing
a set of rooted splits $\mathcal{C}$ that are jointly consistent
and correspond to branches of the input trees. We construct an ordered
list of splits $S(T_{i})$ by walking the branches of tree $T_{i}$
in post-order and appending the corresponding splits to the list.
These lists combined to form the list $\Sigma=S(T_{1})+S(T_{2})+\ldots+S(T_{n})+S(\mathbb{T})$,
where '$+$' denotes concatenation. The set $C$ is initialized to
the empty set, and we then iteratively consider each split in the
list $\Sigma$ and insert it into $C$ if the resulting set remains
jointly compatible. The BUILD algorithm is used both to check compatibility,
and to construct a compatible tree from the final set $C$.

@@

\subsection{Notation}

We extend this framework by adding a set $\mathcal{I}\subseteq\mathcal{N}$
of labels that have the \emph{incertae sedis }property. The taxonomy
consists of the labels $\mathcal{N}$, the taxonomy tree $\taxonomy$,
and the \emph{incertae sedis} property $\mathcal{I}$ of taxon names.
